---
title: "STATS 767 Project Proposal"
author: "Stephen Su, ID 253222842"
output:
  pdf_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
  css: style.css
bibliography: refs.bib
csl: apa.csl
urlcolor: blue
---

<style type="text/css">

body {
  font-family: serif;
  text-align: justify;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r warning=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(janitor)
library(knitr)

# readr::write_csv(
#   readr::read_delim(
#     "https://raw.github.com/ggeop/Myopia-Study/master/myopia.csv",
#     delim = ";"
#   ),
#   "../data/myopia.csv"
# )

myopia <- readr::read_csv(
  "../data/myopia.csv",
  col_types = "__ld______ddddd___"
) %>%
  janitor::clean_names() %>%
  dplyr::mutate(across(
    where(is.logical),
    function(x) case_when(x ~ "Yes", !x ~ "No")
  ))
```

## The Data
The dataset is sourced from [ggeop/Myopia-Study](https://github.com/ggeop/Myopia-Study) [@ggeop] for a study of the risk factors of developing myopia in children, which is a subset from the Orinda Longitudinal Study of Myopia (OLSM) collected in 1989-1990 and 2000-2001. The dataset consists of 618 observations and 17 variables. In particular, the project will focus on the following six numeric variables and one categorical variable:

```{r results="asis",echo=FALSE}
knitr::kable(
  data.frame(
    Variable_Name = names(myopia),
    Unit = c("boolean", "years", rep("hours per week", 5)),
    Description = c(
      "Myopia within the first five years of follow up",
      "Age at first visit",
      "Time spent engaging in sports/outdoor activities",
      "Time spent reading for pleasure",
      "Time spent playing video/computer games or working on the computer",
      "Time spent reading or studying for school assignments",
      "Time spent watching television"
    )
  )
)
```

## Question of Interest
The project aims to explore the relationship between the onset of myopia in children and lifestyle risk factors as per the mentioned variables and produce a statistical predictive model.

## Methodologies
The project will utilise the methods from STATS 767. A linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are considered suitable for modelling the multivariate relationship for the purpose of establishing grouping for the characteristics of myopic and non-myopic children. The LDA and QDA models will also be compared to other methods, including PLS-DA and the logistic regression. The statistical methods used in this project are from **base R** [@baseR], the **MASS** [@MASS] and **mixOmics** [@mixOmics] packages.

## Deliverables of the Project
The final deliverables of the project consist of an RMarkdown report and an HTML slideshow for the presentation using **rmarkdown** [@rmd] and **xaringan** [@xar]. All materials required to reproduce this project can be found at the Github repository [szmsu2011/stats767proj](https://github.com/szmsu2011/stats767proj).


\newpage

## Bibliography
