---
title: "STATS 767 Project: Exploratory Analysis"
author: "Stephen Su, ID 253222842"
output:
  html_document:
    code_folding: hide
    df_print: paged
    fig_caption: yes
    number_sections: yes
  css: style.css
urlcolor: blue
---

<style type="text/css">

body {
  font-family: serif;
  text-align: justify;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev.args = list(type = "cairo-png"))
```
```{r warning=FALSE,message=FALSE}
library(tidyverse)
library(janitor)
library(GGally)

myopia <- readr::read_csv(
  "../data/myopia.csv",
  col_types = "__ld______ddddd___"
) %>%
  janitor::clean_names() %>%
  dplyr::mutate(across(
    where(is.logical),
    function(x) case_when(x ~ "Yes", !x ~ "No")
  ))
```

# The Data
```{r}
myopia

summary(myopia)
```

# Visual Analysis
```{r fig.align='center',message=FALSE}
myopia %>%
  GGally::ggpairs(
    ggplot2::aes(col = myopic),
    columns = 2:7
  )
```
All numeric variables except for `age` are heavily right-skewed.

## Transformation for Mitigating Skewness
```{r fig.align='center',message=FALSE,warning=FALSE}
myopia %>%
  dplyr::mutate(
    log_age = log(age),
    across(-(1:2), log1p, .names = "log1p_{col}"), # log1p since min(x) == 0
  ) %>%
  dplyr::select(-(2:7)) %>%
  GGally::ggpairs(
    ggplot2::aes(col = myopic),
    columns = 2:7
  ) +
  ggplot2::labs(
    title = "Transformed Data",
    subtitle = expression(
      "By" ~ x %->% log(1 + x) ~
      "for all" ~ x ~
      "except for" ~
      age %->% log(age)
    )
  )
```
Logging seems to over-fix skewness of `sporthr` and `tvhr`, no difference for `age`.
```{r fig.align='center',message=FALSE}
(myopia_2 <- myopia %>%
  dplyr::mutate(
    across(4:6, log1p, .names = "log1p_{col}"), # log1p since min(x) == 0
    across(c(3, 7), sqrt, .names = "sqrt_{col}") # Avoid over-fixing
  ) %>%
  dplyr::select(-(3:7))) %>%
  GGally::ggpairs(
    ggplot2::aes(col = myopic),
    columns = 2:7
  ) +
  ggplot2::labs(
    title = "Transformed Data",
    subtitle = expression(
      "By" ~ x %->% log(1 + x) ~
      "for readhr, comphr, studyhr and" ~
      x %->% sqrt(x) ~
      "for sporthr, tvhr"
    )
  )

round(diag(solve(cor(myopia_2[-1]))), 2)
```
The transformed data looks far better, despite substantial departure from normality. The variance inflation factors suggest no multicollinearity.

## Exploration for Effects to Myopia
```{r fig.align='center'}
myopia_2 %>%
  tidyr::pivot_longer(-1, "var") %>%
  ggplot(aes(x = myopic, y = value)) +
  geom_boxplot(outlier.shape = 1) +
  ggplot2::facet_wrap(~var, scales = "free_y") +
  ggplot2::labs(
    title = "Bivariate Association with Myopia",
    x = "", y = ""
  )
```
The bivariate correlations with myopia of the numeric variables are relatively small.
