---
title: "STATS 767 Project"
subtitle: "A Multivariate Approach to Modelling Lifestyle Risk Factors of Children Myopia in the US"
author: "Stephen Su"
date: "2 June 2021"
output:
  xaringan::moon_reader:
    nature:
      highlightLines: true
    css: xaringan-themer.css
---

```{r setup, echo = FALSE, cache = FALSE, results = 'hide'}
library(knitr)
options(
  htmltools.dir.version = FALSE, htmltools.preserve.raw = FALSE,
  tibble.width = 60, tibble.print_min = 6
)
opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, comment = "#>", fig.retina = 3,
  fig.align = "center", fig.show = "hold", dpi = 110,
  dev.args = list(png = list(type = "cairo"))
)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#035AA6",
  header_font_google = google_font("Josefin Sans"),
  text_font_google = google_font("Montserrat", "400", "400i"),
  code_font_google = google_font("Fira Mono")
)
```

```{r external, include = FALSE}
read_chunk("slide.R")
```

## Background

The association of lifestyle factors and myopia has been long discussed.

Since my childhood, I was repeated advised by optometrists that:

* My prescription could keep increasing until adulthood

* Spend more time outdoors, it will save your vision

* **Stay away from those electronic garbage (from Mom)**

* Take regular breaks and avoid prolonged focused reading

With Statistics, can we find supporting evidence for the claims above?

```{r start, echo = 2, error = TRUE}
```

Finding the answer is not easy, and the realm of medical-related statistics is full of variables with random behaviour (weak correlation, confounding, ...)

---

## The Data

The dataset is from [ggeop/Myopia-Study](https://github.com/ggeop/Myopia-Study).

* A subset of the research from Orinda Longitudinal Study of Myopia

* Conducted in 1989-1990 and 2000-2001

* Consists of 618 observations and 17 variables

However, we are only interested in seven of the variables.

```{r data, echo = 4}
```

---

## Variable Definition

```{r data-def, echo = FALSE}
```

---

## A First Glance

```{r prelim, echo = FALSE, fig.height = 4}
```

---

## Log-Transformation But ...

```{r minimum, highlight.output = c(2, 9)}
```
